The Countercyclical Component of Capital Regulation

### 2020-07-17 Analysis/Draft/Draft_bookdown.Rmd
Draft version of first revision by supervisors

---
title: "The Countercyclical Component of Capital Regulation"
date: '\emph{\today}'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    latex_engine: pdflatex
    template: tex-ms.tex
  pdf_document: default
  word_document: default
author:
- name: "Lucas Avezum"
  affiliation: "CentER and EBC, Tilburg University"
biblio-style: apsr
bibliography: ../../Auxiliar/mybibfile.bib
citecolor: black
csl: ../../Auxiliar/cambridge-journal-of-economics.csl
#anonymous: no
fontsize: 11pt
geometry: margin= 1in
header-includes: \usepackage{rotating, graphicx, dcolumn, longtable, float, siunitx, tabularx}
keywords: Internal ratings-based models, Risk-weighted assets dispersion, Procyclicality
linkcolor: black
abstract: The Basel II accord introduced a concave mapping from probabilities of default
  (PD) to capital requirements for banks adopting the internal ratings-based (IRB)
  approach. Therefore, total capital requirements depends not only on the portfolio's
  average PD but also on the shape of its PD distribution. Using a hand-collected
  dataset, I propose a method that decomposes the contribution of the PD distribution
  to capital requirements into an average component and a shape component. While a
  larger average component increases capital charges, a larger shape component generates
  capital savings. I investigate the relationship of these capital savings with the business cycle and find it to   be countercyclical. The results suggest that a recalibration of the current IRB framework can reduce considerably the procyclicality of capital regulation.
spacing: double
subtitle: ''
thanks: Preliminary draft.
JEL: G21, G28, G11
urlcolor: black
toc: FALSE
---

@@

library(tidyverse)            ## Data manipulation, pipe operator
library(data.table)           ## Command to bind lists
library(ggridges)             ## Density ridges plot
library(stargazer)            ## Latex tables
library(lfe)                  ## High-dimensional fixed effects
library(gridExtra)            ## Create grid for plots and tables
library(cowplot)              ## Create grid for plots and tables
library(DescTools)            ## Command to winsorize variables
library(ggpubr)               ## More plots
library(splitstackshape)      ## Command to expand data
library(latex2exp)            ## Use latex code in plots
library(lemon)                ## Plot brackets
#library(textablr)
#library(sandwich)
#library(lmtest)
#library(relaimpo)            ## Calculate relative importance measures

@@

The regulation on bank capital requirements has evolved substantially since the first international standard, the 1988 Basel Accord. At the heart of these changes is the link of capital charges to asset risk [@basel2004international; @BCBS2017basel; @EBA2019policy]. Risk sensitivity of capital requirements was present in the first accord and it was greatly enhanced in the Basel II framework by giving banks the option to calculate capital requirements using risk estimates obtained from their own models, known as the internal ratings-based (IRB) approach. The rationale for risk-linked capital charges is that regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk [@koehn1980regulation; @kim1988risk; @gordy2010risk]. The rationale for allowing internal models is the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit].

However, both risk sensitivity and the use of internal models may have adverse consequences. First, linking capital charges to asset risks exacerbate the procyclicality of lending [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical]. As the overall credit quality deteriorates during economic downturns, capital requirements tighten, forcing banks to cut down lending. Second, the inherent flexibility of the IRB approach may allow for differences in capital requirements that do not reflect portfolio risk but rather modelling choices [@le2012revisiting; @mariathasan2014manipulation; @behn2016limits;@berg2017analysis; @ferribank].

This paper contributes to understanding the cyclicality and variability of capital requirements by identifying the effects of an unexplored feature of the IRB framework on capital requirements. Under the IRB approach, banks supply their own risk estimates, such as probability of default (PD), to formulas set by the regulation which determine the amount of capital banks have to hold. Two characteristics of the framework are important for this analysis. First, the mapping from PD estimates to capital requirements is a concave function. Second capital requirements are calculated for each asset separately and then added up to reach total capital charges. Together, these characteristics imply that actual capital requirements are lower than if they were calculated using the average PD to the entire portfolio. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in equal proportions. The second has the same total amount as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD. Because of the concavity of the IRB formula, capital requirements for the second portfolio are greater than the first. More generally, total capital requirements depend not only on banks' portfolio average PD but also on the shape of the PD distribution.

I propose a measure that decomposes the contribution of the PD distribution to capital requirements into an average component and a shape component. The method is straightforward: for each bank-year combination, compare actual capital requirements to the counterfactual case where requirements for the entire portfolio are calculated using the portfolio’s average PD. I label the shape component *Capital savings* because it reduces total capital requirements.

To calculate measures of *Capital savings*, information on the distribution of banks' regulatory risk parameters is needed. I hand-collected information on the distribution of credit risk parameters from banks' Pillar-3 reports. As is common in the literature, I calculate *Capital savings* in terms of risk-weighted assets (RWA), which are simply capital requirements times 12.5, or risk-weights (RW), which are RWA divided by assets. The sample consists of 59 large banking groups that have adopted the IRB approach. The dataset covers 15 countries and goes from 2007 to 2018. Data at such granularity is not available in the common balance sheet datasets, and although very detailed, datasets from credit registers are restricted to one country, and may still lack the relevant information.

With this novel dataset, I investigate the relationship between *Capital savings* and the business cycle. Prior evidence suggests that capital requirements are procyclical: during economic downturns banks tend to deleverage instead of raising capital. However, as credit conditions deteriorate (and banks react to these new credit conditions) the PD distribution and, consequently, the amount of *Capital savings*, also change. A priori, the direction of the business cycle effect on *Capital savings* is ambiguous. At one extreme, if the entire portfolio is equally affected, then there is only a shift of the location parameter of the PD distribution. Because the PD-RW formula elasticity is lower for higher PD estimates, in this scenario *Capital savings* are procyclical. On the other hand, if riskier assets are more sensitive to credit shocks, then capital savings are countercyclical. The ambiguity arises because *Capital savings* can only increase during a recession if the shape of the PD distribution is affected during the cycle.

I find *Capital savings* to be countercyclical. I estimate that a 1.6% GDP growth rate (one standard deviation) decreases *Capital savings* by 0.7 percentage points on average. In other words, banks have to increase capital by 0.7 percentage points to keep their capital ratio constant. The effect is also economically significant. In a capital ratio decomposition exercise, I also find that the the cyclicality of *Capital savings* reduces the procyclicality of capital ratios by 23.7%. The results are robust to different measures of the shape component, such as, *Capital savings* in terms of ratios, *Capital savings* in monetary units, and using Gini coefficients.

I further analyse the relationship of the PD distribution moments with GDP growth and find that during economic expansions average PD and PD variance decrease while the PD skewness increase. This evidence implies that distributions become more symmetric and with less extreme values during recessions, suggesting a flight to safety behavior by banks. As the recession reduces their existing portfolio credit quality (evidenced by the negative coefficient of average PD) banks respond by reducing risk-taking.

This papers contributes to two strands of the banking literature. First, several studies document significant RW variability across otherwise similar banks. Some take this evidence as indicative of either excessive subjectivity in the current capital requirement rules or risk measurement manipulation by banks to decrease capital charges [see, @mariathasan2014manipulation; @ferribank]. Others are more cautious and emphasize that part of RW variability is desirable since it reflects differences in risk among banks [@cannata2012inside; @BCBS2016regulatory;@berg2017analysis; @EBA2019result]. I contribute to the literature by showing that *Capital savings* are an important source of the RW variation across banks.

@@

The remainder of this paper is organized as follows. Section \ref{sec:background describes the capital requirement formula under the IRB approach, why it creates *Capital savings*, and the hypothesis concerning the cyclicality of *Capital savings*. Section \ref{sec:data} describes the collected data and the measures of *Capital savings*. Section \ref{sec:cycle} presents evidence on the countercyclical feature of *Capital savings*. Section \ref{sec:reallocation} tests if *Capital savings* are affected by portfolio reallocation during the business cycle. Section \ref{sec:additional_results} presentes additional results. Finally, section \ref{sec:conclusion} concludes.

@@

In June 2004, the Basel Committee issued a revised framework on international convergence of capital measurements and capital standards [@basel2004international], known as Basel II, which serves as the basis for national rulemaking and implementation processes. At the heart of the revision was the link of capital charges to asset risk. Since the first accord from 1988 (Basel I), minimum capital requirements are calculated as a percentage of risk-weighted assets. Hence, the approach to define risk-weights is of major importance. In Basel I, risk-weights were flat within portfolio categories. For instance, all sovereign exposures were weighted by 0% and all corporate exposures by 100%. However, regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk as the cost in terms of capital requirements is the same within a portfolio category [@koehn1980regulation; @kim1988risk; @gordy2010risk]. Therefore, since Basel II, financial institutions may choose between two approaches to calculate capital requirements for credit risk; the standardized approach (essentially a slightly modified version of the first accord) and the IRB approach.

In the IRB approach, capital requirements are derived from risk-weight formulas that take as input banks' own estimates of credit risk parameters. The rationale for allowing internal models is the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit]. The basic regulatory formula derives from a model characterized by the property that the risk-weight of each asset depends only on the estimated credit risk parameters for this asset and not on the composition of the portfolio, i.e. the model is portfolio invariant. This feature leads to a bottom-up approach where capital requirements are determined on the asset level and total requirement is simply the sum of asset individual requirement. Equation \ref{eq:RW_function} is the risk-weight formula at its most simple form:

\begin{equation}
\label{eq:RW_function}
\text{RW}(\text{LGD}_i, \text{PD}_i)= 12.5\text{LGD}_i\bigg[ N\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\text{PD}_i)+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg)-\text{PD}_i\bigg],
\end{equation}
where $i$ is an asset in the portfolio and $N$ is the standard normal cumulative distribution. Basel accords segment credit portfolios into several categories (for instance, corporate, sovereign or retail), each having different additional features (maturity adjustments, risk-dependent correlation coefficient, among others) while keeping the basic structure of equation \ref{eq:RW_function}. Because of the portfolio invariant feature of the model, average risk-weight $\text{RW}^f$ is the exposure weighted average of all individual assets' risk-weights, as shown in equation \ref{eq:RW_factual}:

\begin{equation}
\label{eq:RW_factual}
\text{RW}^f= \frac{\sum_{i}\text{EAD}_i \text{RW}(\text{LGD}_i, \text{PD}_i)}{\sum_{i}\text{EAD}_i}= \sum_{i}\text{q}_i \text{RW}(\text{LGD}_i, \text{PD}_i).
\end{equation}
where $q_i= EAD_i/\sum_jEAD_j$. Thus, to calculated their capital requirements banks estimate three credit risk parameters: the exposure at default (EAD), the loss given default (LGD), and the probability of default (PD).[^1] The total amount of RWA for this portfolio is $\text{RW}^f\times \sum_i\text{EAD}_i$ and total capital requirements $\text{RWA}\times 8\%$.

[^1]: For wholesale exposures (corporate, sovereign and banks) the effective maturity of the asset is another parameter and for SMEs annual sales turnover is used to adjust the correlation coefficient.

@@

Figure \ref{fig:pdmap} plots the mapping from PD to RW according to equation \ref{eq:RW_function}. For this paper, the important characteristic of this mapping is that it is a concave function. The concavity of the PD-RW mapping together with the bottom-up method to aggregate capital requirements entail that actual capital requirements are lower than if they were calculated using the average PD to the entire portfolio. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in proportions $q$ and $1-q$. The second has the same total amount as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD weighted by their respective shares. Because of the concavity of the IRB formula and Jensen’s inequality, the capital requirement for the second portfolio is greater than the first. Figure \ref{fig:pdmap} illustrates the comparison. $\text{RW}^f$ and $\text{RW}^c$ are the average RW for the first portfolio and the second portfolio, respectively. More generally, total capital requirements do not depend only on the portfolio average PD but also on the shape of the PD distribution.

I propose a method that decomposes the contribution of the PD distribution to capital requirements into the average component and the shape component. The method is a generalization of the example in Figure \ref{fig:pdmap}: for each bank-year combination, I compare the actual capital requirement ($\text{RW}^f$) to the counterfactual case where the requirement for the entire portfolio is calculated using the portfolio’s average PD. The counterfactual RW is the average component of the PD distribution to capital requirements:

\begin{equation}
\label{eq:RW_counterfactual}
\text{RW}^c= \sum_{i}\text{q}_i \text{RW}(\text{LGD}_i,\overline{\text{PD}}),
\end{equation}
where the PD parameter for all assets is replaced by the portfolio weighted average. The latter is defined as the following:

@@

The difference between $\text{RW}^c$ and $\text{RW}^f$ measures any contribution to capital requirements from the PD distribution that is no related to the location of this distribution, i.e., it is a measure of the shape component. It can also be interpreted as the amount of *Capital savings* per unit of EAD due to the concavity of the IRB formula and the shape of the PD distribution. Equation \ref{eq:RW_savings} shows one of the possibilities to measure *Capital savings*:

@@

Note that I defined *Capital savings* such that positive values translate to lower actual capital requirements and consequently $\text{RW}^s \geq 0$.

@@

Overall, capital requirements under the IRB approach are procyclical. During economic downturns credit risk increases which raises capital requirements forcing banks to either raise equity or cut down lending. Theory and empirical evidence suggest that banks opt for the latter as raising equity is also more expensive during recessions [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical].

Between the two capital requirement components, the average component is a source of the procyclicality as long the relationship between the business cycle and PD is monotonic. Consider again the example shown in Figure \ref{fig:pdmap} with a portfolio containing a safe asset with PD equal to $PD_L$ and a risky asset with PD equal to $PD_H$ in proportions $q$ and $1-q$. In Appendix \ref{sec:appendixa}, I show that for $\text{PD} \in (0,0.25)$, $\text{RW}'>0$, $\text{RW}''<0$ and $\text{RW}'''>0$. I also show in \ref{sec:appendixa} that, on average, XX% of assets' PD lies within this range. For simplicity, I assume a constant $\text{LGD}_i=\text{LGD}$. So, the average component for this portfolio is:

\begin{equation}
\label{eq:RW_counter_example}
\text{RW}^{c}(q,\text{PD}_L, \text{PD}_H)=\text{RW}(q\text{PD}_L+(1-q)\text{PD}_H).
\end{equation}

Because $\text{RW}'>0$ and as long PD estimates are affected monotonic and negatively by GDP growth, the average component of capital requirements increases during economic downturns.

The direction of *Capital savings* during the business cycle is not trivial even when assuming a monotonic relationship between PD estimates and the business cycle. To explain the ambiguity, I consider three cases. But first, note that using the example in Figure \ref{fig:pdmap}, *Capital savings* can be written as the following:

\begin{equation}
\label{eq:RW_savings_example}
\text{RW}^{s}(q,\text{PD}_L, \text{PD}_H)=\text{RW}(q\text{PD}_L+(1-q)\text{PD}_H)-q\text{RW}(\text{PD}_L)+(1-q)\text{RW}(\text{PD}_H).
\end{equation}

We can find the effect of changes in the PD distribution to *Capital savings* by taking the total derivative of \ref{eq:RW_savings}:

>>>>

---
title: "The Countercyclical Component of Capital Regulation"
date: '\emph{\today}'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    #citation_package: natbib
    latex_engine: pdflatex
    template: tex-ms.tex
  pdf_document: default
  word_document: default
author:
- name: "Lucas Avezum"
  affiliation: "CentER and EBC, Tilburg University"
biblio-style: apsr
bibliography: ../../Auxiliar/mybibfile.bib
citecolor: black
csl: ../../Auxiliar/Citation_styles/american-marketing-association.csl
#anonymous: no
fontsize: 11pt
geometry: margin= 1in
header-includes: \usepackage{rotating, graphicx, dcolumn, longtable, float, siunitx, tabularx, appendix} 
keywords: Internal ratings-based models, Risk-weighted assets dispersion, Procyclicality
linkcolor: black
abstract: The Basel II accord introduced a concave mapping from probabilities of default
  (PD) to capital requirements for banks adopting the internal ratings-based (IRB)
  approach. Therefore, total capital requirements depends not only on the portfolio's
  average PD but also on the shape of its PD distribution. Using a hand-collected
  dataset, I propose a method that decomposes the contribution of the PD distribution
  to capital requirements into an average component and a shape component. While a
  larger average component increases capital charges, a larger shape component generates
  capital savings. I investigate the relationship of these capital savings with the business cycle and find it to   be countercyclical. The results suggest that a recalibration of the current IRB framework can reduce considerably the procyclicality of capital regulation.
spacing: double
indent: true
subtitle: ''
thanks: Preliminary draft.
JEL: G21, G28, G11
urlcolor: black
toc: FALSE
---

@@

library(tidyverse)            ## Data manipulation, pipe operator
library(data.table)           ## Command to bind lists
library(ggridges)             ## Density ridges plot
library(stargazer)            ## Latex tables
library(lfe)                  ## High-dimensional fixed effects
library(gridExtra)            ## Create grid for plots and tables
library(cowplot)              ## Create grid for plots and tables
library(multipanelfigure)     ## Create grid for figures
library(DescTools)            ## Command to winsorize variables
library(ggpubr)               ## More plots
library(splitstackshape)      ## Command to expand data
library(latex2exp)            ## Use latex code in plots
library(lemon)                ## Plot brackets
library(mosaic)               ## Package to calculate derivatives
#library(textablr)
#library(sandwich)
#library(lmtest)
#library(relaimpo)            ## Calculate relative importance measures

@@

The regulation on bank capital requirements has evolved substantially since the first international standard, the 1988 Basel Accord. At the heart of these changes is the link of capital charges to asset risk [@BCBS2006international; @BCBS2017basel; @EBA2019policy]. Risk sensitivity of capital requirements was present in the first accord and it was greatly enhanced in the Basel II framework by giving banks the option to calculate capital requirements using risk estimates obtained from their own models, known as the internal ratings-based (IRB) approach. The rationale for risk-linked capital charges is that regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk [@koehn1980regulation; @kim1988risk; @gordy2010risk]. The rationale for allowing internal models is the better alignment of risk estimates to actual portfolio risk [@barakova2014banks] and the promotion of better risk management practices [@cucinelli2018credit].

However, both risk sensitivity and the use of internal models may have adverse consequences. First, linking capital charges to asset risk exacerbate the procyclicality of lending [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical]. As the overall credit quality deteriorates during economic downturns, capital requirements tighten, forcing banks to reduce lending. Second, the inherent flexibility of the IRB approach may allow for differences in capital requirements across banks that do not reflect portfolio risk but rather modelling choices [@le2012revisiting; @mariathasan2014manipulation; @behn2016limits;@berg2017analysis; @ferribank].

This paper contributes to understanding the cyclicality and variability of capital requirements by identifying the effects of an unexplored feature of the IRB framework on capital requirements. Under the IRB approach, banks supply their own risk estimates, such as the probability of default (PD), to formulas set by the regulation which determine the amount of capital banks have to hold. Two characteristics of the framework are important for this analysis. First, the mapping from PD estimates to capital requirements is a concave function. Second, capital requirements are calculated for each asset separately and then added up to obtain the total capital charge. Because of these characteristics, actual capital requirements for a portfolio are lower than implied by the average PD of this portfolio. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in equal proportions. The second has the same total amount as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD. Because of the concavity of the IRB formula, capital requirements for the second portfolio are greater than capital requirements for the first. More generally, total capital requirements depend not only on banks' portfolio average PD but also on the shape of the PD distribution.

I propose a measure that decomposes the contribution of the PD distribution to capital requirements into an average component and a shape component. The average component is a counterfactual scenario such that the capital requirements for the entire portfolio are calculated using the portfolio’s average PD. The shape component is obtained as the difference between the average component and the actual amount of capital requirements. I label the shape component *Capital savings* because an increase of the shape component reduces total capital requirements.

To calculate *Capital savings*, I hand-collected information on the distribution of credit risk parameters from banks' Pillar-3 reports. The sample consists of 59 large banking groups that have adopted the IRB approach. The dataset covers 15 countries and goes from 2007 to 2018.

With this novel dataset, I investigate the relationship between *Capital savings* and the business cycle. Between the two capital requirement components, the average component is a source of the procyclicality as long the relationship between the business cycle and PD estimates is monotonic. Conversely, the direction of *Capital savings* during the business cycle is not trivial even when assuming a monotonic relationship between PD estimates and the business cycle. The ambiguity arises because *Capital savings* can only increase during a recession, and therefore be countercyclical, if the shape of the PD distribution is affected during the business cycle. If the PD estimates of the the assets in a portfolio are equally affected during the business cycle, then there is only a shift of the location parameter of the PD distribution. Because the PD elasticity of capital requirements is lower for higher PD estimates, in this scenario *Capital savings* are procyclical. On the other hand, if riskier assets are sufficiently more sensitive to credit shocks than safer assets, then *Capital savings* are countercyclical.

I find *Capital savings* to be countercyclical. I estimate that a 1.7% GDP growth rate (one standard deviation) decreases the growth rate of *Capital savings* by 0.8 percentage points on average. In other words, given an average PD, banks have to increase capital by 0.8 percentage points to keep their capital ratio constant. The effect is also economically significant. The variation of *Capital savings* during the business cycle reduces the procyclicality of capital ratio by 23.7% on average. The results also suggest that the countercyclical relationship between GDP and *Capital savings* is mostly driven by wholesale portfolios, especially corporate portfolios. The results are robust to different measures of the shape component, to the inclusion of several control variables and the alternative sample selections. I also show that selection into the IRB approach does not explain the cyclicality of *Capital savings*.

I further analyse the relationship of the PD distribution moments with GDP growth and find that during economic recessions average and variance of PD distributions increase while the their skewness decrease. Although the effect of the average PD is stronger, which characterizes capital requirements as procyclical, both the higher variance and the lower skewness of the PD distribution mitigate the increase of capital requirements during recessions. Moreover, I argue that the change of skewness during the business cycle is consistent with a portfolio reallocation effect on *Capital savings*.

This papers contributes to two strands of the banking literature. First, several studies document significant capital requirements variability across otherwise similar banks. Some take this evidence as indicative of either excessive subjectivity in the current capital requirement rules or risk measurement manipulation by banks to decrease capital charges [see, @mariathasan2014manipulation; @ferribank]. Others are more cautious and emphasize that part of the capital requirements variability is desirable since it reflects differences in risk among banks [@cannata2012inside; @BCBS2016regulatory;@berg2017analysis; @EBA2019result]. I contribute to the literature by showing that *Capital savings* are an important source of capital requirements variation across banks.

@@

The remainder of this paper is organized as follows. Section \ref{sec:background} describes the capital requirement formula under the IRB approach, why it creates *Capital savings*, and the hypothesis concerning the cyclicality of *Capital savings*. Section \ref{sec:data} describes the collected data and the measures of *Capital savings*. Section \ref{sec:cycle} presents evidence on the countercyclical feature of *Capital savings*. Section \ref{sec:reallocation} tests if *Capital savings* are affected by portfolio reallocation during the business cycle. Section \ref{sec:additional_results} presents additional results. Finally, section \ref{sec:conclusion} concludes.

@@

In June 2004, the Basel Committee issued a revised framework on international convergence of capital measurements and capital standards [@BCBS2006international], known as Basel II, which serves as the basis for national rulemaking and implementation processes. The link of capital charges to asset risk was at the heart of the revision. Since the first accord from 1988 (Basel I), minimum capital requirements are calculated as a percentage of risk-weighted assets (RWAs). Hence, choosing the approach that defines risk-weights is crucial. In Basel I, RWs were constant within portfolio categories. For instance, all sovereign exposures were weighted by 0% and all corporate exposures by 100%. However, regulating capital via a simple capital to asset ratio can incentivize banks to take on excessive risk as the cost in terms of capital requirements is the same within a portfolio category [@koehn1980regulation; @kim1988risk; @gordy2010risk]. Therefore, since Basel II, financial institutions may choose between two approaches to calculate capital requirements for credit risk; the standardized approach (essentially a slightly modified version of the first accord) and the IRB approach.

Under the IRB approach, banks are required to estimate their credit risk parameters using internal risk models. The risk parameters are the probability of default (PD), the loss given default (LGD), and the exposure at default (EAD). The main rationale for allowing internal models is the better alignment of regulatory risk estimates to actual portfolio risk [@barakova2014banks] as banks are expected to have better knowledge of and more resources to monitor their portfolio than supervisors have. Because the internal models have to be approved by the supervisors before implemented for regulatory purposes, regulators also expected the IRB approach to promote stronger risk management practices in the banking industry [@BCBS2006international; @cucinelli2018credit].

The estimated risk parameters are then used in regulatory formulas for calculating risk-weights. The regulatory formulas comes from a model characterized by the property that the risk-weight of each asset depends only on the estimated credit risk parameters for this asset and not on the composition of the portfolio, i.e. the model is portfolio invariant. This feature leads to a bottom-up approach where capital requirements are determined on the asset level and total requirement is simply the sum of asset individual requirement. Equation \ref{eq:RW_function} is the risk-weight formula at its most simple form:

\begin{equation}
\label{eq:RW_function}
\text{RW}(\text{LGD}_i, \text{PD}_i)= 12.5\text{LGD}_i\bigg[ N\bigg(\sqrt{\frac{1}{1-R}}N^{-1}(\text{PD}_i)+\sqrt{\frac{R}{1-R}}N^{-1}(0.999)\bigg)-\text{PD}_i\bigg],
\end{equation}
where the subscript $i$ indexes an asset in the portfolio, $N$ is the standard normal cumulative distribution, and $R$ is a correlation coefficient. Basel accords segment credit portfolios into several categories each having different additional features while keeping the basic structure of equation \ref{eq:RW_function}. For instance, the correlation coefficient is a function of PD for wholesale exposures while it is constant at 15% for real estate exposures.[^1] Although the regulatory formulas are common for all banks the scope that banks are allowed to use their internal models varies. Under the foundation IRB (F-IRB) approach, banks can apply or be approved to use only internal PD estimates and are required to use values defined by the regulator for the other parameters. Only banks approved to the advanced IRB (A-IRB) approach can use internal estimates for all the risk parameters.

Because of the portfolio invariant feature of the model, average risk-weight, $\text{RW}^f$, is the exposure weighted average of all individual assets' risk-weights, as shown in equation \ref{eq:RW_factual}:[^2]

\begin{equation}
\label{eq:RW_factual}
\text{RW}^f= \sum_{i}q_i \text{RW}(\text{LGD}_i, \text{PD}_i).
\end{equation}
where $q_i= \text{EAD}_i/\sum_j\text{EAD}_j$. The total amount of RWAs for this portfolio is $\text{RW}^f\times \sum_i\text{EAD}_i$ and total capital requirements $\text{RWA}\times 8\%$.

[^1]: Some other differences are maturity adjustments for wholesale exposures and the use of annual sales turnover to adjust the correlation coefficient for SMEs.

[^2]: The superscript $f$ (for factual) is used to differentiate the average value from the RW function.

@@

Figure \ref{fig:pdmap} plots the mapping from PD to risk-weights according to equation \ref{eq:RW_function}. For the purpose of this paper, the important characteristic of this mapping is that it is a concave function. The concavity of the regulatory formula together with the bottom-up method to aggregate capital requirements entail that total capital requirements for a portfolio do not depend only on the portfolio average PD but also on the shape of the PD distribution. For clarity, consider two portfolios. The first contains a low PD asset and a high PD asset in proportions $q$ and $1-q$. The second has the same amount of exposure as the first but contains only a mean PD asset, where mean PD is the mean of low PD and high PD weighted by their respective shares. Because of the concavity of the IRB formula and Jensen’s inequality, the capital requirement for the second portfolio is greater than the first. Figure \ref{fig:pdmap} illustrates the comparison. $\text{RW}^f$ and $\text{RW}^c$ are the average risk-weight for the first portfolio and the second portfolio, respectively. More generally, any PD distribution with positive variance is charged with capital requirements lower than implied by the average of the distribution.   
I propose a method that decomposes the contribution of the PD distribution to capital requirements into the average component and the shape component. The method is a generalization of the example in Figure \ref{fig:pdmap}. The average component is the amount of capital requirements if the portfolio’s average PD is used for the entire portfolio. Similar to $\text{RW}^c$ in Figure \ref{fig:pdmap}, the average component is defined as the following:

\begin{equation}
\label{eq:RW_counterfactual}
\text{RW}^c= \sum_{i}q_i \text{RW}(\text{LGD}_i,\overline{\text{PD}}),
\end{equation}
where the PD parameter for all assets is replaced by the portfolio weighted average. The latter is defined as the following:

@@

The difference between $\text{RW}^c$ and $\text{RW}^f$ measures any contribution to capital requirements from the PD distribution that is not related to the location of this distribution, i.e., it is a measure of the shape component. It can also be interpreted as the amount of *Capital savings* per unit of EAD due to the concavity of the IRB formula and the shape of the PD distribution. I define this intensity measure of *Capital savings* as the following:

@@

The total amount of *Capital savings* is simply $\text{RW}^{s}$ multiplied by total EAD in the portfolio as follows:

\begin{equation}
\label{eq:measure2}
\text{RWA}^{\$}=\text{RWA}^c-\text{RWA}^f=\text{RW}^{s}\times \sum_i\text{EAD}_i.
\end{equation}

Because this is a measure in monetary units, $\text{RWA}^{\$}$ is mechanically larger for larger banks and they increase as the portfolio rolls out to the IRB approach.

An alternative way to measure the intensity of *Capital savings* is the ratio of equation \ref{eq:RW_counterfactual} to equation \ref{eq:RW_factual} as follows:

\begin{equation}
\label{eq:measure1}
\text{RWA}^s=\frac{\text{RWA}^c}{\text{RWA}^f}.
\end{equation}

$\text{RWA}^s$ is the rate at which a bank saves capital. For instance, a value of 1.5 means this portfolio would require 50% more capital in the absence of *Capital savings* to keep the same capital ratio. A value of one means that there are no *Capital savings* and implies a portfolio with a degenerate PD distribution. Any PD distribution with positive variance results in positive *Capital savings*.

$\text{RWA}^s$ is also my preferred measure of *Capital savings* because it is the most straightforward measure to evaluate the impact of *Capital savings* on the capital ratio during the business cycle. In the empirical analysis, I study the behavior of variables during the cycle by relating their growth rates to GDP growth rate. Hence, I regress the change in the logarithm of my variables of interest on GDP growth rate. Taking my empirical approach into account, $\text{RWA}^s$ becomes the natural measure of *Capital savings* as the following decomposition of the log of capital ratio shows:

\begin{equation}
\begin{aligned}
\label{eq:decomposition}
\text{Log}\bigg(\frac{\text{Capital}}{\text{RWA}^f}\bigg) &= \text{Log Capital} -  \text{Log RWA}^f \\
&= \text{Log Capital} -  \text{Log}\bigg(\frac{\text{RWA}^c}{\text{RWA}^s}\bigg) \\
&= \text{Log Capital} -  \text{Log RWA}^c + \text{Log RWA}^s
\end{aligned}
\end{equation}

If I regress all the change of these four variables - $\frac{\text{Capital}}{\text{RWA}^f}$, $\text{Capital}$, $\text{Log RWA}^c$, and $\text{Log RWA}^s$ - on GDP growth rate I obtain four estimates $\hat\beta_{\text{Cap. ratio}}$, $\hat\beta_{\text{Capital}}$, $\hat\beta_{\text{RWA}^c}$, and $\hat\beta_{\text{RWA}^s}$, respectively. Using the predicted values implied by these estimated in Equation \ref{eq:decomposition} we get the following  equality: $\hat\beta_{\text{Cap. ratio}} = \hat\beta_{\text{Capital}} -\hat\beta_{\text{RWA}^c} + \hat\beta_{\text{RWA}^s}$, which can be used to obtain the relative importance of changes of each of these three variables to changes of the capital ratio during the business cycle.

Note that I defined *Capital savings* such that positive values translate to lower actual capital requirements and consequently $\text{RW}^s \geq 0$, $\text{RWA}^{\$} \geq 0$, and $\text{RWA}^s \geq 1$.

Lastly, for robustness of the empirical analysis, I also calculate the Gini coefficient of the PD distribution as an alternative measure to my approach.

@@

Overall, capital requirements under the IRB approach are procyclical. During economic downturns credit risk increases which raises capital requirements forcing banks to either raise equity or cut down lending. Theory and empirical evidence suggest that banks opt for the latter as raising equity is costly, even more so during recessions [@danielsson2001academic; @kashyap2004cyclical; @repullo2012procyclical; @behn2016procyclical].

Between the two capital requirement components, the average component is a source of the procyclicality as long the relationship between the business cycle and PD is negative and monotonic. Consider again the example shown in Figure \ref{fig:pdmap} with a portfolio containing a share $q$ of a safe asset with PD equal to $\text{PD}_L$ and a share $1-q$ of a risky asset with PD equal to $\text{PD}_H$ such that $\text{PD}_H>\text{PD}_L$. For practical reasons I restrict the analysis to the  $\text{PD} \in [0,0.25)$. Figure \ref{fig:densities} in Appendix \ref{sec:appendixa} shows that the percentage of PD within the $[0,0.25)$ interval is between 99.35% and 99.98%. Figure \ref{fig:derivatives} in Appendix \ref{sec:appendixa} shows that for $\text{PD} \in [0,0.25)$, $\frac{\partial\text{RW}}{\partial\text{PD}}>0$, $\frac{\partial^2\text{RW}}{\partial\text{PD}^2}<0$, and $\frac{\partial^3\text{RW}}{\partial\text{PD}^3}>0$. I assume, without loss of generality, a constant $\text{LGD}_i=\text{LGD}$ and constant $q$. Using the weighted average of this portfolio in equation \ref{eq:RW_counterfactual}, the average component for this portfolio is equal to the following:

\begin{equation}
\label{eq:RW_counter_example}
\text{RW}^{c}(\text{PD}_L, \text{PD}_H)=\text{RW}(q\text{PD}_L+(1-q)\text{PD}_H).
\end{equation}

Because $\frac{\partial\text{RW}}{\partial\text{PD}}>0$ and as long PD estimates are affected monotonic and negatively by GDP growth, the average component of capital requirements increases during economic downturns.

Conversely, the direction of *Capital savings* during the business cycle is not trivial even when assuming a monotonic relationship between PD estimates and the business cycle.  The ambiguity arises because *Capital savings* can only increase during a recession, and therefore be countercyclical, if the shape of the PD distribution is affected during the business cycle. If the PD estimates are equally affected during the business cycle, then there is only a shift of the location parameter of the PD distribution. Because the PD elasticity of capital requirements is lower for higher PD estimates, in this scenario *Capital savings* are procyclical. On the other hand, if riskier assets are sufficiently more sensitive to credit shocks than safer assets, then *Capital savings* are countercyclical. To illustrate the ambiguity, I consider three cases.

First, note that using the example in Figure \ref{fig:pdmap}, *Capital savings* can be written as the following:

\begin{equation}
\label{eq:RW_savings_example}
\text{RW}^{s}(\text{PD}_L, \text{PD}_H)=\text{RW}(q\text{PD}_L+(1-q)\text{PD}_H)-q\text{RW}(\text{PD}_L)+(1-q)\text{RW}(\text{PD}_H).
\end{equation}

We can find the effect of changes in the PD distribution to *Capital savings* by taking the total derivative of \ref{eq:RW_savings_example}:
